### 2025-12-18 Paper/paper_2025.tex
Updates from Overleaf

	\linespread{1.2}\title{\vspace{-0.5in} Unobserved Firm Affiliations: Evidence from Hospital Board Overlap}

@@

    Antitrust enforcement in the US has been particularly relevant in recent years as price gouging and consolidation has increased.

    Antitrust legislation creates an additional hurdle for large-scale anti-competitive behavior, but there are other types of firm affiliations that typically elude antitrust scrutiny such as common ownership, joint ventures, and other types of affiliation/partnerships. By and large, firms form partnerships with other firms, often competitors, without needing to seek approval. This has started to raise concern recently, and states have even begun regulating smaller-scale affiliations in recent years. Horizontal mergers and consolidation have rightfully received a lot of attention in the literature, often linked to something?? However, we know little about the effects of smaller-scale affiliations, which could carry some of the same implications as consolidation proposals scrutinized for their ant-competitive nature.

    In this paper, I examine the effects of soft consolidation in the form of overlapping board members in nonprofit hospitals. Existing research establishes a strong link between board overlap and governance practices, but we know little about whether there should be anti-competitive concerns when firms share common board members. Additionally, most studies have been limited to large, publicly traded firms. This focus stems from the assumption that nonprofits engage in less anti-competitive behavior (\cite{baer2014clayton}; \cite{aai2013section7}), making board overlap in such organizations less concerning. However, this assumption is flawed, particularly in industries like health care, where nonprofit hospitals dominate and have been shown to engage in anti-competitive practices (\cite{hulver2023ftc}). Despite these insights, gaps persist: research rarely addresses market-level effects or nonprofit sectors, assuming nonprofits pose minimal antitrust risk (\cite{baer2014clayton}; \cite{aai2013section7}). This assumption is problematic in health care, where nonprofit hospitals dominate and exhibit anti-competitive behavior (\cite{hulver2023ftc}). Further work should explore competitive implications of board overlap beyond governance and financial metrics, particularly in nonprofit contexts.

    In 2024, the FTC banned a case of interlocking board (\cite{ftc2025sevita}).

    \section{Background/Setting}

@@

    \section{Data and Summary Statistics}

@@

    \subsection{Documenting Overlapping Board Members}

@@

    add outcome graphs

@@

>>>>

	\linespread{1.2}\title{\vspace{-0.5in} Non-Merger Affiliations in Already Consolidated Health Care Markets}

@@

    Market concentration has increased across industries in the last two decades (\cite{grullon2019us}), sparking renewed debate and, more recently, a shift toward stricter antitrust enforcement. Modern antitrust policy primarily focuses on mergers and acquisitions among large firms. Since the Hart–Scott–Rodino Act of 1976, the U.S. government has reviewed certain transactions that may threaten competition, defined by regulators as “a process of rivalry that incentivizes businesses to offer lower prices, improve wages and working conditions, enhance quality and resiliency, innovate, and expand choice” (\cite{DOJFTC2023MergerGuidelines}). Formal challenges remain rare: since 2015, roughly 3\% of proposed mergers have faced extended review, and 32 proposals were dissolved or restructured (\cite{ConversableEconomist2025Antitrust}).

    Careful scrutiny of large-scale mergers is important. However, growing empirical and policy concern centers on unreported, small-scale mergers and non-merger affiliations. Research shows that small-scale mergers, often called “stealth mergers”, can reduce competition (\cite{majerovitz2021consolidation}; \cite{aggarwal2025stealth}; \cite{wollmann2019stealth}; \cite{wollmann2020get}; \cite{kepler2023stealth}). Non-merger affiliations, by contrast, encompass structural ties such as joint ventures, management consolidation, and overlapping boards of directors that do not trigger merger review. Among these, board interlocks are particularly noteworthy because they create direct governance links between firms, are explicitly addressed under the Clayton Act, and have drawn increasing policy attention. While board interlocks are documented (\cite{LemleyVanLoo2025OverlappingDirectors}; \cite{Poberejsky2025InterlocksInnovation}), their competitive effects remain largely hypothesized.

    This issue is especially relevant in health care, where consolidation is a persistent concern and several states have begun regulating small-scale mergers and non-merger affiliations (\cite{ncsl2024healthcare}). Additionally, in September 2025, the FTC announced the resignation of three directors from Sevita Health after uncovering board overlap with Beacon Specialized Living Services, both providers of residential care for individuals with intellectual and developmental disabilities, citing concerns under the Clayton Act and urging firms to proactively avoid such conflicts (\cite{ftc2025sevita}). Board interlocks are no longer abstract governance issues but active antitrust targets. Thus, in this paper, I construct a measure of non-merger affiliation among U.S. nonprofit hospitals that captures both interlocking boards and executive cross-board service, and I examine how these affiliations influence hospital behavior.

    I collect data on nonprofit hospital executives and board members from publicly available IRS Form 990 filings and merge these records with the American Hospital Association (AHA) Annual Survey. Using this combined dataset, I classify hospital affiliations based on system membership and geographic market. I define a non-merger affiliation as the presence of a common board member or an executive serving on another hospital’s board, where the two hospitals are otherwise unaffiliated. That is, they do not belong to the same system or network but share leadership ties. Because the hypothesized effects of non-merger affiliations are anti-competitive, whether the affiliated hospitals compete for the same patients is a critical distinction. To capture this, I separate affiliations into those occurring within the same Hospital Referral Region (HRR) and those across different HRRs, where HRRs represent geographic hospital markets (\cite{Wennberg1973SmallAV}). I show that [X]\% of nonprofit hospitals in my sample share a leader with an otherwise unaffiliated hospital in the same market, and [Y]\% of HRRs contain at least one such affiliation. I consider several hospital outcomes as behaviors that are informed by the corporate governance literature and hospital consolidation literature. I group these outcomes into categories of investment, patient composition, service offerings, and operating costs.

    Similarly to the merger literature, selection into the decision to become affiliated complicates causal inference. I begin with a staggered difference-in-differences design comparing hospitals that form an affiliation to those that do not, and I show that [INSERT FINDINGS]. While event-study estimates reveal no evidence of differential pre-trends, unobserved factors may still influence affiliation decisions. Therefore, my preferred specification employs a triple-differences approach that leverages variation across time, affiliation status, and whether the affiliation occurs within the same geographic market. This design isolates the incremental effect of affiliating with a hospital in the same market where competitive effects are most salient. Using this approach, I find that [INSERT FINDINGS].

    [IMPLICATIONS OF FINDINGS] Firms are responsive to antitrust regulation. That is, enforcement is a deterrant for anti-competitive behavior. Thus, it is important to think about whether regulation should increase even for small-scale affiliations. (something like this for policy implications?)

    Prior work establishes that corporations are deeply interconnected through overlapping directors and that these ties shape governance practices and firm performance (\cite{dooley1969interlocking}; \cite{ferris2003too}; \cite{field2013busy}; \cite{bizjak2009option}; \cite{bouwman2011corporate}; \cite{chiu2013board}; \cite{cai2014board}; \cite{baran2017director}; \cite{geng2021does}; \cite{schonlau2009board}; \cite{cai2012board}; \cite{cheng2021social}). That literature largely centers on publicly traded firms and frames overlap as either a governance friction (busy directors, option backdating, strategy convergence) or a source of private benefits (director networks that raise value, profitability, and post‑M\&A performance). In contrast, I study non‑merger affiliations in nonprofit hospitals, an arena often presumed less prone to anti-competitive behavior (\cite{baer2014clayton}; \cite{aai2013section7}) despite mounting evidence of competitive risks in health care (\cite{hulver2023ftc}). I broaden the construct beyond traditional board interlock to include executive cross‑board service, and I move the focus from governance proxies and firm‑level valuation to market‑relevant behaviors that are informed by the literature on both governance and consolidation (\cite{baran2017director}; \cite{geng2021does}; \cite{schonlau2009board}; \cite{cai2012board}; \cite{cheng2021social}). Methodologically, I leverage a triple‑differences design that distinguishes affiliations formed within versus across geographic markets to isolate the incremental, competition‑salient effect of overlap, rather than inferring behavior from financial outcomes alone.

    A large empirical literature examines the effects of consolidation in health care, particularly through hospital mergers and system integration. Early work found little evidence that consolidation improved operating costs, efficiency, or quality of care (\cite{alexander1996short}; \cite{ho2000hospital}; \cite{dranove2003hospital}). Yet from 2000 to 2020, hospital mergers surged, leaving a quarter of markets with no independent hospitals by 2020 (\cite{ElevanceHealth2023}). This trend spurred extensive research on price, quality, and efficiency. Studies consistently document modest cost and efficiency gains for hospitals joining larger systems (\cite{schmitt2017hospital}; \cite{andreyeva2024corporatization}; \cite{craig2021mergers}), but these gains rarely translate into lower prices; instead, mergers typically lead to significant price increases as hospitals gain bargaining power with insurers (\cite{gaynor2012impact}; \cite{boozary2019association}; \cite{cooper2019price}; \cite{andreyeva2024corporatization}). Evidence on quality is mixed: most studies find little effect on readmissions or mortality (\cite{haas2011mergers}; \cite{beaulieu2020changes}), though some report increased treatment intensity and mortality for heart disease patients (\cite{hayford2012impact}) or higher readmission rates overall (\cite{andreyeva2024corporatization}). Beyond price and quality, research highlights mechanisms such as convergence in treatment styles (\cite{eliason2020acquisitions}), consolidated service offerings (\cite{mariani2022impact}), and redistribution of patients within markets, including declines in Medicaid admissions in highly concentrated areas (\cite{desai2023hospital}). While this literature focuses on mergers and system affiliation, my study examines a different form of structural connection: non-merger affiliations through board and executive overlap.

    The paper proceeds as follows. In Section \ref{sec:background}, I provide institutional details on nonprofit hospital governance and hospital affiliations. In Section \ref{sec:data}, I outline the data collection process and provide summary statistics. In Section \ref{sec:simpledid}, I estimate a simple staggered difference-in-differences model leveraging affiliations over time. Finally, in Section \ref{sec:tripledid}, I leverage variation in both affiliation and geography to estimate a triple differences specification.

    \section{Background/Setting}\label{sec:background}

@@

    \section{Data and Summary Statistics}\label{sec:data}

@@

    \subsection{Documenting Leadership Overlap}

@@

    \import{Objects}{hosp_outcomes_table.tex}

    \section{Is Leadership Overlap Correlated with Consolidation?}

    While hospitals often publicly declare that their leadership affiliations are not intended to lead to mergers or other  consolidation (add citations), the relationship between leadership overlap and consolidation is an open empirical question.    Theoretically, there are three ways that leadership overlap and consolidation could be related. First, leadership overlap may act as a substitute for formal mergers. In this case, hospitals leverage shared governance to achieve some benefits of consolidation, such as strategic alignment, resource coordination, and cost efficiencies, without incurring the regulatory scrutiny and transaction costs associated with a merger. If this hypothesis holds, we would expect leadership overlap to reduce the likelihood of subsequent merger activity. Second, leadership overlap may serve as a precursor to merging. Hospitals might begin by sharing board members or executives to build trust, harmonize decision-making, and prepare for operational integration. Under this interpretation, overlap is an intermediate step toward full consolidation, and its presence should increase the probability of a future merger. Third, leadership overlap may result from intentions to merge. Here, overlap is not an independent strategic choice but rather a signal of an already-decided merger plan. This scenario introduces endogeneity concerns, as leadership overlap and merger intentions are jointly determined. In addition to its possible role in mergers, leadership overlap may also influence the likelihood of hospital closure. Shared leadership could provide access to broader networks, financial resources, and managerial expertise, which might help struggling hospitals remain viable and avoid closure. Conversely, overlap could signal strategic realignment within a system, where weaker hospitals are deliberately phased out to concentrate resources in stronger facilities. In this sense, leadership overlap could either reduce closure risk by stabilizing operations or increase it if consolidation strategies involve selective shutdowns.

    To evaluate these possibilities, I estimate the effect of gaining or losing leadership overlap on two outcomes: (1) the likelihood that a hospital closes in the future, and (2) the likelihood that it merges in the future. I employ a stacked difference-in-differences (DiD) approach, which constructs event-time observations for each treated hospital around the time of gaining or losing overlap and stacks these into a single panel. This method accommodates variation in treatment timing across hospitals and enables estimation of dynamic effects.

    \begin{equation}
    Y_{it} = \alpha + \sum_{k \neq -1} \beta_k \cdot D_{i,t+k} + \gamma_i + \delta_t + \varepsilon_{it}
    \end{equation}

    \noindent where $Y_{it}$ is the outcome for hospital $i$ at time $t$ (closure or merger indicator), $D_{i,t+k}$ is an indicator for event time $k$ relative to the change in leadership overlap (with $k = -1$ as the omitted category), $\gamma_i$ and $\delta_t$ are hospital and time fixed effects, respectively, and $\beta_k$ captures the effect at each event time $k$.

    \begin{figure}[ht!]
        \centering
        \vspace{9mm}
        \caption{Effect of Overlap on Closure}
        \includegraphics[width=.9\textwidth]{Objects/closure.pdf}
        \label{fig:closure}
    \end{figure}

    \begin{figure}[ht!]
        \centering
        \vspace{9mm}
        \caption{Effect of Overlap on in-System Overlap}
        \includegraphics[width=.9\textwidth]{Objects/any_formal_sameHRR.pdf}
        \label{fig:any_formal_sameHRR}
    \end{figure}

    \begin{figure}[ht!]
        \centering
        \vspace{9mm}
        \caption{Effect of Overlap on Being Independent}
        \includegraphics[width=.9\textwidth]{Objects/independent.pdf}
        \label{fig:independent}
    \end{figure}

@@

    \appendix

\begin{figure}
    \centering
    \caption{Outcome Variables Over Time, Patient and Service Allocation}
    \includegraphics[width=0.75\linewidth]{Objects/outcome_perc_dis_bed_graph.pdf}
    \label{fig:outcome_means1}
\end{figure}

\begin{figure}
    \centering
    \caption{Outcome Variables Over Time, Purhcases, Costs, and Revenue}
    \includegraphics[width=0.75\linewidth]{Objects/outcome_purch_oper_rev_graph.pdf}
    \label{fig:outcome_means2}
\end{figure}
